Quantity discrimination in the common garden skink and the role of early-life conditions

### 2025-03-02 ms.qmd
updates

---
title: "Do early environmental experiences impact numerical discrimination abilities in a lizard?"
docx: 
  reference-doc: "./bib/tmpl.docx"
bibliography: "./bib/refs.bib"
csl: "./bib/behavioral-ecology-and-sociobiology.csl"
execute:
  echo: false
  error: false
  cache: false
  warning: false
link-citations: true
crossref:  
  fig-title: Fig    # (default is "Figure")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Table  # (default is "Table")
editor_options: 
  chunk_output_type: console
---

@@

@@

## Abstract

**From social behaviour to navigating complex environments, quantitative abilities can be crucial to fitness. However, early life conditions can impact brain development to affect quantitative competence. For example, early thermal conditions or glucocorticoid concentrations (GCs) – stress-related hormones transmitted by the parents – can play a prominent role in shaping cognition through their effects on brain physiology. Furthermore, temperature can lead to increased levels of GCs that can be transmitted to offspring, emphasizing the need to study the combined effects of these two factors. Here, we investigated the effects of elevated prenatal corticosterone (CORT) – the main GC in reptiles – and incubation temperature on numerical discrimination in the common garden skink (_Lampropholis guichenoti_). We assessed numerical abilities through a spontaneous choice test with food as a stimulus. Employing a repeated measures design, we subjected lizards to five numerical tests each differing in the ratios between the two choices (1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4). Contrary to our predictions, we found no evidence for treatment effects on lizard behaviour and no use of numerical discrimination during foraging in _L. guichenoti_. Our findings build on previous studies suggesting that reptiles feeding on live prey may have difficulty distinguishing between different prey quantities. Despite our results, further investigation of numerical abilities in live-prey-feeding reptiles and the impact of early conditions on other taxa is warranted, as these areas remain understudied.**

## Significance statement

**Discriminating between quantities can be crucial for individual fitness, yet the impact of early environmental factors on the development of quantitative competence remains largely unexplored. In this study, we investigated, for the first time, the effects of prenatal corticosterone exposure and incubation temperature on numerical discrimination in a reptile. Contrary to our predictions, we did not observe any evidence of numerical discrimination and therefore no impact of early developmental environment. Our results align with previous studies showing that reptiles consuming live prey struggle to differentiate prey numbers in spontaneous choice tests. These results open new avenues for exploring the evolution of numerical abilities in reptiles and identifying factors influencing the development of quantitative skills in this group. Further research should explore alternative methods, such as trained tasks, to better understand the cognitive mechanisms underlying quantitative competence in reptiles.**

@@

The main two approaches experimental approaches in the study of quantity discrimination are spontaneous choice tests and training procedures. In spontaneous choice tests, the subjects are presented two groups of biologically relevant stimuli differing in some quantitative trait (number, size, volumen, etc) and the experimenters record if the subjects select or not the most most appropriate quantity (e.g. the group with more conspecifics or the larger amount of food) (@agrillo2014spontaneous; @cox2016quantity). In training procedures, animals must learn a numerical rule to receive a reward [@agrillo2014spontaneous; @miletto2018quantity]. Both methods are believed to be complementary to each other, and both can provide different results even in the same species [see @miletto2017quantitative and @miletto2018quantity]. One reason is that spontaneous choice tests assume that one of the options presented is more adequate than the other when this may not always be the case. For example, choosing groups of conspecifics with less individuals could be advantageous when food is scarce but disadvantageous when predators are abundant [**REFs first thesis??**]. As such, context and motivation can be important drivers of decision-making that can change the way we measure animals' abilities to discriminate between quantities [@agrillo2014spontaneous]. The other reason is that each method may tap into different cognitive mechanisms [@agrillo2014spontaneous]. In this sense, **Researchers have suggested that numerousness can be processed by two types of systems: one system for small sets (≤ 4) that works by tracking and encoding objects as individual entities; and an analogous system that operates with large numerosities by estimation processes (for a review, see Cantlon et al. (2009); Hyde (2011).**   **HOW CAN EACH BE RELATED TO MOTIVATION OR LEARNING??**

**Differences in test discrete (OFS) versus overall quantity/size (ANS). How to measure both? Creating NUMBER inconsistencies with the SAME SIZE (examples).** OFS, can be challenging, since it is necessary to control for other continuous variables such as area, density, or movement (@agrillo2014spontaneous; **Mehlis et al. 2015**). Generally, authors increase one variable at a time without increases in the other to give conflicting information about the correlation between numerical cues and quantitative continuous traits [**REFs**]. For example, **Look for something where they increase number controlling for size** **HOW??**

No other amphibian or non-avian reptile study has set out to
manipulate these variables to a degree as thoroughly as Stancher et al.42. Future experiments
involving amphibians and non-avian reptiles that manipulate these variables will be crucial for
understanding the relationship between the core systems and the information available for
discrimination

These studies have been traditionally focused on mammals, birds, and fishes despite the influence of numerical cognition on fitness across taxa [@nieder2005counting; @agrillo2014spontaneous]. However, recent studies demonstrate that other vertebrates like amphibians or non-avian reptiles are also competent in numbers [@uller2003salamanders; @stancher2015numerical; @miletto2017quantitative]. For example, red-footed turtles (_Chelonoidis carbonaria_) rapidly learned to associate colour with higher volumes of food [@soldati2017long], Hermann’s tortoises (_Testudo hermanni_) discriminate between large and small food quantities in a spontaneous choice test [@gazzola2018continuous], and freshwater turtles (_Trachemys scripta_ and _Mauremys sinensis_) rapidly learned a quantity discrimination task [@lin2021superior; @lin2024trained]. The Italian wall lizard (_Podarcis siculus_) failed to discriminate between different numbers of food items in a spontaneous choice test [@miletto2017quantitative], but succeeded in a trained quantity discrimination task [@miletto2018quantity]. In contrast, Iberian rock lizards (_Iberolacerta cyreni_) spontaneously preferred larger quantities of crickets when presented five versus ten items [@recio_prey_2021]. **In skinks,**

Lizards were presented with food quantities
differing in number or size and were able to select the larger quantity of food items when they differed in number but not when items
differed in size. We show, for the first time, superior spontaneous discrimination of items differing in number over size in a lizard spe-
cies, which contrasts with previous findings. Our simple method, however, did not include controls for the use of continuous quantities,
and further tests are required to determine the role of such information during quantity discrimination. [@szabo_spontaneous_2021]

We subjected lizards to a Y-maze experiment presenting both discrete (comparisons involving two
vs. four, two vs. six, two vs. eight, four vs. six, four vs. eight, and six vs. eight pieces) and equivalent continuous combinations (single
pieces differing in size) of pumpkin pieces. Our findings revealed that sleepy lizards exhibited spontaneous quantity discrimination, fa-
voring the larger quantity in three discrete comparisons (eight vs. two, two vs. six, four vs. six) and two continuous comparisons (eight
vs. two, eight vs. four). However, low accuracy may be attributable to difficulty in visual acuity exacerbated by difficulty discriminating
quantities at small ratios. [@szabo2024spontaneous]

**These studies demonstrate that reptiles have cognitive abilities related to quantity discrimination that have been overlooked in the past, but that should be explored further.**

Cognitive abilities, however, can be strongly influenced by developmental conditions when the brain is particularly sensitive to environmental inputs [@zhu_prenatal_2004]. For instance, the temperature at which embryos develop can affect neuron density or learning abilities in different species of reptiles [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @dayananda_incubation_2017; @siviter_incubation_2017; @abayarathna_effects_2020]. These thermal effects on cognition may arise directly or indirectly through impacts on parental condition [@Crino_2023]. Temperatures outside the optimal range can trigger stress responses, leading to the release of glucocorticoids (GCs) [@sapolsky_how_2000], which may then be transmitted to the offspring [@Crino_2023]. Elevated prenatal GCs can, in turn, influence cognition by altering brain function [@szuran_water_1994; @zhu_prenatal_2004; @crino_corticosterone_2014-learn; @farrell_developmental_2015-learn; @bebus_associative_2016; @lui2017chronic]. As such, the interaction between GCs and develpomental temperature must be relevant in shaping reptiles' cognition. Most studies on the effects of GC exposure or elevated temperatures during development focused on learning and memory, but few studies have investigated how early-life conditions can affect quantitative abilities despite the importance of numerical discrimination [but see @vila_pouca_quantity_2019]. Furthermore, the interactive effects of prenatal GCs and temperature on cognitive abilities remain largely unexplored.

Here, we investigated the use of numerical cues in the common garden skink (_Lampropholis guichenoti_) using a spontaneous choice test where lizards were presented with different number of food items while **controlling** for total length. In adition, we examined how prenatal corticosterone (CORT) - the main GC in reptiles - and incubation temperature can affect numerical discrimination and decision-making in foraging contexts. To do so, we manipulated CORT concentration (CORT-treated or a sham control) in _L. guichenoti_ eggs and incubated them under one of two temperature regimes (Cold - 23 ± 3ºC or Hot - 28 ± 3ºC) in a 2x2 factorial design. Then, we assessed numerical discrimination abilities on five tests that differed in the relative number of food items: 1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4 crickets.  Previous studies in other skinks showed that these species can discriminater between different quantities of food, although the cues employed to do so remain unclear [@szabo_spontaneous_2021; @szabo2024spontaneous]. We predicted _L. guichenoti_ to be able to discriminate between different number of crickets even when controlling by total length. Furthermore, we predicted lizards' decisions to be influenced by the early conditions they were subjected to. In this regard, we hypothesized that lizards incubated at higher temperatures would outperform those incubated at lower temperatures in the numerical discrimination task [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @vila_pouca_quantity_2019]. We predicted that increased CORT levels would impair numerical competence [@szuran_water_1994; @zhu_prenatal_2004], with more pronounced effects at lower temperatures. Consistent with Weber's law, we predicted that lizards incubated at higher temperatures would successfully discriminate between quantities even at high ratios (e.g. 3 VS 4), whereas even the lowest ratios (e.g. 1 VS 4) could be challenging for those exposed to CORT or incubated at lower temperatures.

@@

*Hatchlings* -- Immediately after hatching, we measured Snout-Vent Length (SVL) and Tail Length (TL) with a ruler to the nearest mm (± 0.1 mm), and mass using a digital scale (OHAUS, Model spx123) to the nearest g (± 0.001 g). Hatchlings were then placed in individual enclosures (18.7L x 13.2W x 6.3H cm) provided with non-stick matting and a small water dish. During this period, they were given water daily and received 3-6 small *A. domestica* crickets three times a week. All care otherwise follows similar protocols to adults (see above).

@@

The numerical discrimination test consisted of a series of two-choice trials where lizards were presented with different numbers of food items. Each choice was placed on one of the two lateral platforms of a white 3D printed device (see @fig-Methods B, C) and then covered with a transparent Petri dish to prevent the lizard from obtaining the food. We performed five tests that differed in the ratio between both choices of crickets. Specifically, we compared: 1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4 crickets. The food consisted of small frozen crickets (*A. domestica*) that were thawed and dusted with calcium and multivitamins. We decided to use frozen crickets to avoid movement that would affect lizard behaviour [@cooper2024tell].

Lizards were fed frozen crickets for one month prior to the experiments to habituate them to eating frozen prey. To habituate lizards to the experimental conditions, frozen crickets were placed on top of one of the platforms and Petri dishes during each feeding for two weeks prior to experimental trials. The side where the food was placed was changed every day to avoid any side bias during the trials. However, to further control for side biases, the higher number of crickets was randomly placed on the right or the left platform. We also counterbalanced the number of individuals per treatment tested in the same type of trial each day.

The crickets used during the experiment were all the same size, but to control for other quantitative traits we changed the orientation of the crickets as illustrated in @fig-Methods D. This approach created a conflict of information between choices, as horizontally oriented crickets could occupy the same space as a greater number of vertically oriented crickets. This allowed us to determine whether the lizards relied on discrete numerical units rather than non-numerical cues, such as total length, area, or volume [@agrillo2014spontaneous]. In addition, after the experiments, we conducted another set of trials to test whether the orientation of the crickets could affect lizards’ choice, but we found no effect of orientation (see _Supplementary Material - Prey orientation test_).

@@

All tests were done between 1000-1200 hours when the lizards were most active. Each trial was performed daily for five days between the 17^th^ and 22^nd^ of May 2024. We recorded three main variables: i) _Latency_, defined as the time between the start of the test - when the platform was left in the enclosure - until the first interaction with one of the options; ii) First choice of the lizard (_Choice_), recorded as 1 if the lizard interacted first with the larger amount or 0 if not; and iii) _Interest_ shown for the higher amount of food (detailed below). We considered that the lizard interacted with the food when it touched the Petri dish with its snout or when the front half of the body was on top of the platform. For five minutes after the first interaction with one of the options, we recorded the time spent interacting with that option for a maximum of five minutes; we did this for both amounts of food independently of what was the first choice. We measured the interest shown for the higher amount of food (_Interest_) by subtracting the time interacting with the option with fewer crickets to the time spent interacting with the larger amount of food. If there was no interaction with one of the options during the test, we considered the time spent as zero for that option. When there was no interaction with any of the choices, both _Latency_, _Choice_, and _Interest_ were noted as NA. We also considered Interest as NA in a few cases where the animals managed to lift the dish and eat the prey; Latency and Choice were recorded and employed in the analyses. We excluded from the analyses all the individuals that did not make any choice in three or more tests. All the analyses were performed by PR, who was blind to the treatment of the lizards.

@@

#### Foraging context may influence prey number discrimination in reptiles  
Lack of numerical discrimination in _L. guichenoti_ may be due to insectivorous lizards relying on other cues, such as movement or size, for foraging decisions. In studies where reptiles discriminated between different amounts of food, the relevant stimuli involved either vegetables [@gazzola2018continuous; @szabo_spontaneous_2021] or large quantities (e.g., 5 vs. 10)[@recio_prey_2021]. In contrast, when _Podarcis siculus_ were presented different numbers of _Musca domestica_ larvae in a spontaneous choice test, the lizards failed to discriminate between quantities despite showing high quantitative skills when prey size was the main cue [@miletto2017quantitative] or when tested using training procedures [@miletto2018quantity]. Notably, the numbers of food items employed in @miletto2017quantitative were always fewer than four. Spontaneous choice tests assume that animals should select the most appropriate quantity, which may not necessarily be the largest one [@agrillo2014spontaneous]. This occurrs if the costs of selecting the larger option outweigh the benefits. For example, when presented two groups with different numbers of conspecifics or if prey handling costs are related to larger quantitative traits like size [@agrillo2014spontaneous]. It could be that the perceived handling costs of larger numbers of prey affected lizard response in our study. Nevertheless, in that scenario, we would still expect animals to exhibit a preference, even if for the smaller amount of food. Additionally we would have expected body size to influence lizards choices, as larger animals may be more adept at handling a greater number of prey items. @miletto2017quantitative did not report any effect of body size, and age - a proxy for size - did not affect _L. guichenoti_'s _Choice_ or _Interest_ in our study. The absence of preference in _P. siculus_ and _L. guichenoti_ suggests that the benefits of selecting a larger number of prey are not different from selecting fewer. For instance, when considering hunting groups of 1 VS 4 or 2 VS 4 crickets, the probability of success might be similar. Furthermore, while one individual handles a prey item within the patch, other conspecifics could capture the rest. As a result, insectivorous lizards might not have evolved a preference for larger patches of prey, at least with small numbers [see @recio_prey_2021].

Conversely, our findings may be influenced by the experimental design. In our study, we rotated the position of the crickets in the choice with fewer items to create a situation where the numerical cues (fewer items) conflicted with other quantitative cues associated with numerosity. This approach was intended to control for other quantitative cues that could be used by lizards to make decisions [@cooper2024tell]. However, this manipulation could have made the task more challenging for lizards, potentially affecting their performance. Indeed, it is possible that _L. guichenoti_ uses multiple quantitative cues simultaneously during foraging decisions. When those cues conflict, the preference for one cue over another might vary between individuals, potentially producing the results observed here. Future studies should explore the use of other quantitative cues in _L. guichenoti_ and how they interact with numerical information. Such studies would enhance our understanding of how _L. guichenoti_ integrates different quantitative cues during foraging, providing insights into their decision-making processes and overall quantitative abilities.

#### Foraging behaviour is robust to early environmental experiences  
Although the effects of early environment on cognition is widely studied, little attention has been paid to its impacts on numerical discrimination. To the best of our knowledge, there is only one study examining how developmental conditions influence numerical abilities [@vila_pouca_quantity_2019]. Given the potent effects of early environment on cognition and brain development [@szuran_water_1994; @zhu_prenatal_2004; @amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @lui2017chronic], it is crucial to explore how early environmental conditions shape numerical abilities across taxa.

>>>>

---
title: "Quantity discrimination in the common garden skink and the role of early-life conditions"
docx: 
  reference-doc: "./bib/tmpl.docx"
bibliography: "./bib/refs.bib"
csl: "./bib/behavioral-ecology-and-sociobiology.csl"
execute:
  echo: false
  error: false
  cache: false
  warning: false
link-citations: true
crossref:  
  fig-title: Fig    # (default is "Figure")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Table  # (default is "Table")
editor_options: 
  chunk_output_type: console
---

@@

```{r, data_length_trials}
#| label: data_length_trials
#
source(here("R", "func.R"))
#
data_length <- read.csv(here("./data/measures_crickets.csv")) %>%
  group_by(test) %>%
  summarize(
    mean_length_small = format_dec(mean(measure_small, na.rm = TRUE), 2),
    mean_length_large = format_dec(mean(measure_big, na.rm = TRUE), 2),
    mean_difference = format_dec(mean(diff_measure, na.rm = TRUE), 2),
    low_diff = format_dec(quantile(diff_measure, 0.025, na.rm = TRUE), 2),
    high_diff = format_dec(quantile(diff_measure, 0.975, na.rm = TRUE), 2)
  ) %>%
  ungroup() %>%
  mutate(
    `95% CI` = paste0("[", low_diff, " , ", high_diff, "]")
  ) %>%
  dplyr::select(test, mean_length_small, mean_length_large, mean_difference, `95% CI`)
```

@@

## Abstract
From social behavior to navigating complex environments, quantity discrimination essential for fitness. Animals can use one of two systems to discriminate between different quantities, with variation across taxa. In reptiles, quantity discrimination remained largely unexplored until recently, and the use of one system over the other can depend on the species or the context in which they are tested. Additionally, early life conditions can influence brain development and, in turn, quantity discrimination. For example, early thermal conditions or glucocorticoid concentrations (GCs) — stress-related hormones - when transmitted by the mother, can significantly shape cognition through their effects on brain physiology. Here, we investigated numerical abilities in the common garden skink (_Lampropholis guichenoti_), as well as the effects of elevated prenatal GCs and incubation temperature on numerical discrimination. We assessed numerical abilities using a spontaneous choice test with food as a stimulus. Using a repeated measures design, we subjected lizards to five numerical tests (1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4 prey items), while controlling for continuous variables. Contrary to our predictions, we found no evidence of numerical discrimination in _L. guichenoti_. Furthermore, prenatal conditions did not influence their performance or decision-making. Our findings contribute to the growing body of evidence suggesting that reptiles feeding on live prey may struggle to distinguish between quantities in spontaneous choice tests. Despite these results, further investigation into numerical abilities in live-prey-feeding reptiles and the impact of early conditions on other taxa remains an important area of research.

## Significance statement
Discriminating between quantities can be crucial for individual fitness, yet the ability to distinguish quantities remains poorly understood in reptiles. In this study, we investigated numerical discrimination in _Lampropholis guichenoti_, using a spontaneous choice test with food as the stimulus. Contrary to our predictions, we found no evidence of numerical discrimination in this species. In addition, we explored the impact of early environmental factors, specifically prenatal corticosterone exposure and incubation temperature, on numerical competence. However, we did not observe any effect of these early-life conditions on lizard decision-making or performance. These results open new avenues for exploring the evolution of numerical abilities in reptiles and identifying factors influencing the development of quantity discrimination in this group. Further research should explore alternative methods, such as trained tasks, to better understand the cognitive mechanisms underlying numerical competence in reptiles

@@

The main two approaches experimental approaches in the study of quantity discrimination are spontaneous choice tests and training procedures. In spontaneous choice tests, the subjects are presented two groups of biologically relevant stimuli differing in some quantitative traits (number, size, volumen, etc) and the experimenters record if the subjects select or not the most most appropriate quantity (e.g. the group with more conspecifics or the larger amount of food) (@agrillo2014spontaneous; @cox2016quantity). In training procedures, animals must learn a numerical rule to receive a reward [@agrillo2014spontaneous; @miletto2018quantity]. Both methods are believed to be complementary to each other, and both can provide different results even in the same species [see @miletto2017quantitative and @miletto2018quantity]. One reason is that spontaneous choice tests assume that one of the options presented is more adequate than the other when this may not always be the case. For example, choosing groups of conspecifics with less individuals could be advantageous when food is scarce but disadvantageous when predators are abundant [@yee2013costs]. As such, context and motivation can be important drivers of decision-making that can change the way we measure animals' abilities to discriminate between quantities [@agrillo2014spontaneous].

The other reason is that each method may engage distinct cognitive mechanisms when processing numerical information [@agrillo2014spontaneous]. In this regard, researchers propose that two core systems underlie numerical cognition: the approximate number system (ANS) and the object file system (OFS) [@hyde2011two]. The ANS estimates the approximate magnitude of items within a set, with accuracy decreasing as the numerical difference between sets decreases — a principle known as Weber’s law [@hyde2011two]. In contrast, the OFS allows for the precise tracking of discrete items but is thought to be limited to small quantities, typically up to three or four [@hyde2011two]. The engagement of these systems depends not only on the numerical magnitude but also on contextual factors such as the availability of continuous variables (e.g., area, density, movement) and the nature of the task [@stancher2015numerical]. The presence of such continuous cues typically favors ANS-based processing, while in their absence, the OFS is more likely to be employed [@agrillo2014spontaneous].

The other reason is that each method may tap into different cognitive mechanisms [@agrillo2014spontaneous]. In this sense, researchers consider two core systems within the brain that allow for the processing of quantity information: the approximate number system (ANS) and the object file system (OFS) [@hyde2011two]. ANS encodes the approximate magnitudes of items within sets, with accuracy decreasing as the distance between numbers decreases, obeying the called Weber’s law [@hyde2011two]. OFS, on the other hand, allows for the precise tracking of discrete items within sets, and it is thought to be limited to small numbers (typically up to three or four) [@hyde2011two]. However, the use of either system may be dictated, not only by the magnitude of the stimulus, but also by factors such as the availability of continuous information or the context of the task [@stancher2015numerical]. For example, the presence of continuous variables such as area, density, or movement usually entails the use of the ANS, while the OFS would be predominant in the absence of those continuos cues (@agrillo2014spontaneous). This often complicates the study of numerosity, as continuous traits must be carefully controlled to ensure the use of the OFS [@agrillo2014spontaneous]. In spontaneous choice tests, for instance, animals may rely on cumulative surface area rather than numerical differences when selecting between two sets of food items [@agrillo2014spontaneous]. To disentangle these cues, researchers often equate the total area of the sets while changing the number of items, forcing individuals to rely on numerical information alone [@stancher2015numerical].

Quantity discrimination studies on both systems have been traditionally focused on mammals, birds, and fishes despite the influence of numerical cognition on fitness across taxa [@mehlis2015quantification; @nieder2018evolution]. However, recent studies demonstrate that other vertebrates like amphibians or non-avian reptiles are also competent in numbers [@uller2003salamanders; @stancher2015numerical; @miletto2017quantitative]. For example, red-footed turtles (_Chelonoidis carbonaria_) rapidly learned to associate colour with higher volumes of food [@soldati2017long], Hermann’s tortoises (_Testudo hermanni_) discriminate between large and small food quantities in a spontaneous choice test [@gazzola2018continuous], and freshwater turtles (_Trachemys scripta_ and _Mauremys sinensis_) rapidly learned a quantity discrimination task [@lin2021superior; @lin2024trained]. The Italian wall lizard (_Podarcis siculus_) failed to discriminate between different numbers of food items in a spontaneous choice test [@miletto2017quantitative], but succeeded in a trained quantity discrimination task [@miletto2018quantity]. In contrast, Iberian rock lizards (_Iberolacerta cyreni_) spontaneously preferred larger quantities of crickets when presented five versus ten items [@recio_prey_2021]. In skinks, Shinglebacks (_Tiliqua rugosa_) exhibited spontaneous quantity discrimination using both discrete and continuous cues [@szabo2024spontaneous], while Gidgee skinks (_Egernia stokesii_) discriminated between itemst when they differed in number, but not when size was the main quantitative cue [@szabo_spontaneous_2021]. These studies demonstrate that reptiles have cognitive abilities related to quantity discrimination that have been overlooked in the past, and that the use of the ANS versus the OFS can differ between species and in different contexts [@szabo_spontaneous_2021]. However, cognitive performance is not solely dictated by species differences; developmental conditions can also shape an individual's ability to process numerical information [@vila_pouca_quantity_2019].

Cognition can be strongly influenced by developmental conditions because the brain is particularly sensitive to environmental inputs in this life stages [@zhu_prenatal_2004]. For instance, the temperature at which embryos develop can affect neuron density or learning abilities in different species of reptiles [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @dayananda_incubation_2017; @siviter_incubation_2017; @abayarathna_effects_2020]. These thermal effects on cognition may arise directly or indirectly through impacts on parental condition [@Crino_2023]. Temperatures outside the optimal range can trigger stress responses, leading to the release of glucocorticoids (GCs) [@sapolsky_how_2000], which may then be transmitted to the offspring [@Crino_2023]. Elevated prenatal GCs can, in turn, influence cognition by altering brain function [@szuran_water_1994; @zhu_prenatal_2004; @crino_corticosterone_2014-learn; @farrell_developmental_2015-learn; @bebus_associative_2016; @lui2017chronic]. As such, the interaction between GCs and develpomental temperature must be relevant in shaping reptiles' cognition. Most studies on the effects of GC exposure or elevated temperatures during development focused on learning and memory, but few studies have investigated how early-life conditions can affect quantity discrimination despite its importance on animals' lifes [but see @vila_pouca_quantity_2019]. Furthermore, the interactive effects of prenatal GCs and temperature on cognitive abilities remain largely unexplored.

Here, we investigated the use of numerical cues in the common garden skink (_Lampropholis guichenoti_) using a spontaneous choice test where lizards were presented with different number of food items while controlling for continuous variables. In adition, we examined how prenatal corticosterone (CORT) - the main GC in reptiles - and incubation temperature can affect numerical discrimination and decision-making in foraging contexts. To do so, we manipulated CORT concentration (CORT-treated or a sham control) in _L. guichenoti_ eggs and incubated them under one of two temperature regimes (Cold - 23 ± 3ºC or Hot - 28 ± 3ºC) in a 2x2 factorial design. Then, we assessed numerical discrimination abilities on five tests that differed in the relative number of food items: 1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4 crickets.  Previous studies in other skinks showed that these species can discriminater between different quantities of food using both the ANS and the OFS [@szabo_spontaneous_2021; @szabo2024spontaneous]. We predicted _L. guichenoti_ to be able to discriminate between different number of crickets even when controlling by other continuous variables (i.e. by using the OFS). Furthermore, we predicted lizards' decisions to be influenced by the early conditions they were subjected to. In this regard, we hypothesized that lizards incubated at higher temperatures would outperform those incubated at lower temperatures in the numerical discrimination task [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @vila_pouca_quantity_2019]. We predicted that increased CORT levels would impair numerical competence [@szuran_water_1994; @zhu_prenatal_2004], with more pronounced effects at lower temperatures. We predicted that lizards incubated at higher temperatures would successfully discriminate between quantities even at high ratios (e.g. 3 VS 4), whereas even the lowest ratios (e.g. 1 VS 4) could be challenging for those exposed to CORT or incubated at lower temperatures.

@@

*Hatchlings* -- Immediately after hatching, we measured snout-vent length (SVL) and tail length (TL) with a ruler to the nearest mm (± 0.1 mm), and mass using a digital scale (OHAUS, Model spx123) to the nearest g (± 0.001 g). Hatchlings were then placed in individual enclosures (18.7L x 13.2W x 6.3H cm) provided with non-stick matting and a small water dish. During this period, they were given water daily and received 3-6 small *A. domestica* crickets three times a week. All care otherwise follows similar protocols to adults (see above).

@@

The numerical discrimination test consisted of a series of two-choice trials where lizards were presented with different numbers of food items. Each choice was placed on one of the two lateral platforms of a white 3D printed device (see @fig-Methods B, C) and then covered with a transparent Petri dish to prevent the lizard from obtaining the food. We performed five tests that differed in the ratio between both choices of crickets. Specifically, we compared: 1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4 crickets. The food consisted of small frozen crickets (*A. domestica*) that were thawed and dusted with calcium and multivitamins. We decided to use frozen crickets to avoid movement that would affect lizard behaviour [@cooper2024tell]. Lizards were habituated to eating frozen crickets for one month before the experiments.

The crickets used during the experiment were all the same size, but to control for some continuous traits we changed the orientation of the crickets as illustrated in @fig-Methods D. This method was intended to control for length while minimizing increases in convex hull as much as possible (see @stancher2015numerical). We tested whether the area occupied was similar in all the test by measuring the total length occupied by all the crickets in both choices for half of the animals (see _Supplementary Material - Control of size in both options_), but we did not explicitly control for convex hull. Our results showed that except for the 1 VS 4 and 1 VS 3 tests, the difference in length between both choices was not different from zero. As such, discrimination between both choices in the rest of the tests (2 VS 4, 2 VS 3, 3 VS 4 crickets) would be produced by the OFS, not the ANS [@agrillo2014spontaneous]. In addition, after the experiments, we conducted another set of trials to test whether the orientation of the crickets could affect lizards’ choice, but we found no effect of orientation (see _Supplementary Material - Prey orientation test_).

To habituate lizards to the experimental conditions, frozen crickets were placed on top of one of the platforms and Petri dishes during each feeding for two weeks prior to experimental trials. The side where the food was placed was changed every day to avoid any side bias during the trials. However, to further control for side biases, the higher number of crickets was randomly placed on the right or the left platform. We also counterbalanced the number of individuals per treatment tested in the same type of trial each day.

@@

All tests were done between 1000-1200 hours when the lizards were most active. Each trial was performed daily for five days between the 17^th^ and 22^nd^ of May 2024. We recorded three main variables: i) _Latency_, defined as the time between the start of the test - when the platform was left in the enclosure - until the first interaction with one of the options; ii) First choice of the lizard (_Choice_), recorded as 1 if the lizard interacted first with the larger amount or 0 if not; and iii) _Interest_ shown for the higher amount of food (detailed below). We considered that the lizard interacted with the food when it touched the Petri dish with its snout or when the front half of the body was on top of the platform. For five minutes after the first interaction with one of the options, we recorded the time spent interacting with that option for a maximum of five minutes; we did this for both amounts of food independently of what was the first choice. We measured the interest shown for the higher amount of food (_Interest_) by subtracting the time interacting with the option with fewer crickets to the time spent interacting with the larger amount of food. If there was no interaction with one of the options during the test, we considered the time spent as zero for that option. When there was no interaction with any of the choices, both _Latency_, _Choice_, and _Interest_ were noted as NA. We also considered Interest as NA in a few cases where the animals managed to lift the dish and eat the prey; Latency and Choice were recorded and employed in the analyses. We excluded from the analyses all the individuals that did not make any choice in three or more tests. All the analyses were performed by PR, who was blind to the treatment of the lizards, but not the type of test or the cricket's number as the videos were analysed manually.

@@

#### Foraging context may influence prey number discrimination in reptiles  
Our aim in this study was to investigate whether _L. guichenoti_ could discriminate between different number of crickets in a spontaenous choice test relying exclusively on the OFS. This question arose from previous research showing that other species within the Scincidae family could use the OFS in spontaneous discrimination tests [@szabo_spontaneous_2021]. To control for non numerical cues like total length and convex hull, we rotated the position of the crickets in the choice with fewer items, creating a situation where total length was similar in both choices [@agrillo2014spontaneous; @stancher2015numerical; @cooper2024tell]. This design intended to test whether _L. guichenoti_ could rely on numerical information alone when the ANS was not available. However, this manipulation could have made the task more challenging for lizards, potentially affecting their performance. If _L. guichenoti_ primarily relies on the ANS for quantity discrimination, this could explain the absence of a numerical preference. Nonetheless, in two of our tests (1 VS 4 and 1 VS 3 crickets) our manipulation was not sufficient to control for total length (see _Supplementary Material - Control of size in both options_). Yet, we still did not find quantity discrimination in _L. guichenoti_. This suggests that the absence of numerical preferences in our study is unlikely to be an artifact of our manipulation, but rather reflects a genuine lack of quantity discrimination in this species, at least in spontaneous choice tests.

In this regard, lack of numerical discrimination in _L. guichenoti_ may be due to insectivorous lizards relying on other cues, such as movement or size, for foraging decisions. In studies where reptiles discriminated between different amounts of food, the relevant stimuli involved either vegetables [@gazzola2018continuous; @szabo_spontaneous_2021] or large quantities (e.g., 5 vs. 10)[@recio_prey_2021]. In contrast, when _Podarcis siculus_ were presented different numbers of _Musca domestica_ larvae in a spontaneous choice test, the lizards failed to discriminate between quantities despite showing quantity discrimination when prey size was the main cue [@miletto2017quantitative] or when tested using training procedures [@miletto2018quantity]. Notably, the numbers of food items employed in @miletto2017quantitative were always fewer than four. Spontaneous choice tests assume that animals should select the most appropriate quantity, which may not necessarily be the largest one [@agrillo2014spontaneous]. This occurrs if the costs of selecting the larger option outweigh the benefits. For example, when presented two groups with different numbers of conspecifics or if prey handling costs are related to larger quantitative traits like size [@agrillo2014spontaneous]. It could be that the perceived handling costs of larger numbers of prey affected lizard response in our study. Nevertheless, in that scenario, we would still expect animals to exhibit a preference, even if for the smaller amount of food. Additionally we would have expected body size to influence lizards choices, as larger animals may be more adept at handling a greater number of prey items. @miletto2017quantitative did not report any effect of body size, and age - a proxy for size - did not affect _L. guichenoti_'s _Choice_ or _Interest_ in our study. The absence of preference in _P. siculus_ and _L. guichenoti_ suggests that the benefits of selecting a larger number of prey are not different from selecting fewer. For instance, when considering hunting groups of 1 VS 4 or 2 VS 4 crickets, the probability of success might be similar. Furthermore, while one individual handles a prey item within the patch, other conspecifics could capture the rest. As a result, even if diet might not affect quantity discrimination abilities, insectivorous lizards might not have evolved a preference for larger patches of prey, at least with small numbers [see @recio_prey_2021].
